\section*{Acknowledgments}
\label{sec::acknowlegments}

The warmest thanks for this work go to my advisor, Knut Reinert.
He always trusted me, motivated me, showed his interest, let me free to work on my ideas. %and let me look at the bright side of things.
Now I really understand how lucky I have been.

Special thanks go to Raffaele Giancarlo, first for having introduced me to the field of Bioinformatics, then for having referred me to the International Max Planck Research School (IMPRS), and finally for having accepted to review this thesis.
I am also grateful to my IMPRS advisors Martin Vingron and Peter N. Robinson, as well as to the IMPRS coordinator Kirsten Kelleher ---
this work has been possible thanks to the financial support of the IMPRS.

Huge thanks go to David Weese for teaching me his art of computer programming and constantly working with me --- without his help I would not have gone far.
My gratitude goes also to Jochen Singer for jointly working on the FM-index and to Manuel Holtgrewe for the evaluations of Masai.
And thanks to Jacopo Pantaleoni, Jonathan Cohen and Mark Berger for the great experience at NVIDIA!

Thanks --- for too many reasons --- to the IMPRSers Akdes, Alessandro, Anne-Katrin, Arthur, Birte, Chris, Leon, Matt, Mike, Sabrina, Sandro, Stephan, Temesgen, Victor, Xiao.
Thanks to the people at the Free University, including Anja for remembering me to lock the door, Christian for keeping my productivity low, Edzard, Katrin, Ren\'e and Julianus for going lunch late, Jialu for letting me taste wonderful Chinese tea, and the new SeqAn people Hannes, Chenxu, Jongkyu.

Warm thanks for their support go to my parents Anna and Giovanni, my brothers Alessandro and Daniele --- I wish I could spend more time with them -- and Vija, the sweetest thing in my life.
