\chapter{Read mappers parameterization}

\section{Masai evaluation}
\label{sup:masai:param}

In the following, I give the exact parameterization of each read mapper considered in the evaluation of section \ref{masai:evaluation}.

\paragraph{Masai}
Version 0.5 was used.
In order to use Masai as an all-mapper, I passed the argument \texttt{--all}, otherwise the argument \texttt{--any-best} is used by default.
I set the maximal edit distance using the parameter \texttt{-e}.
I configured the seed length with the parameter \texttt{--seed-length}; on \emph{E.~coli}, \emph{D.~melanogaster} and \emph{C.~elegans} I chose a seed length of $16$, while on \emph{H.~sapiens} I chose a seed length of $33$.
I selected the SAM output format with \texttt{-os} and enabled CIGAR output with \texttt{-oc}.

\paragraph{Bowtie\,2}
Version 2.0.0-beta6 was used.
I used the parameter \texttt{--end-to-end} to enforce semi-global read alignments.
For the Rabema experiment I used the parameter \texttt{-k 100}.

\paragraph{BWA}
Version 0.6.1-r104 was used.
For the Rabema experiment I passed the parameter \texttt{-N} to \texttt{aln} and \texttt{-n 100} to \texttt{samse}.

\paragraph{Soap\,2}
Version 2.1 was used.

\paragraph{RazerS\,3}
Version 3.1 was used.
I mapped with indels using the pigeonhole filter (default) and set the error rate through the parameter \texttt{-i}, \eg \texttt{-i 95} to map within an error rate of 5\,\%.
I selected the native or SAM output format with \texttt{-of 0} or \texttt{-of 4}.

\paragraph{Hobbes}
Version 1.3 was used.
I built the index using the recommended $q$-gram length 11.
Since I focus on edit distance, I used the 16\,bit bit-vector version.
I enabled indels with \texttt{--indels} and set maximal edit distance using the parameter \texttt{-v}.
For resource measurement I used the output without CIGAR, for analyzing the results I enabled CIGAR output using \texttt{--cigar}.

\paragraph{mrFAST}
Version 2.1.0.6 was used.
I set maximal edit distance using the parameter \texttt{-e}.

\paragraph{SHRiMP\,2}
Version 2.2.2 was used.


\section{Yara evaluation}
\label{sup:yara:param}

In the following, I give the exact parameterization of each read mapper considered in the evaluation of section \ref{sec:yara:eval}.
Below, \texttt{MIN} and \texttt{MAX} are placeholders for minimal and maximal insert size, while \texttt{INS} is the mean insert size and \texttt{ERR} its allowed deviation, \ie \texttt{INS = (MIN + MAX) / 2}, \texttt{ERR = (MAX - MIN) / 2}.

\paragraph{Yara}
Version 0.9.4 was used.
By default, the tool runs as a best-mapper; I set the number of strata with the parameter \texttt{-s}.
To perform all-mapping, I passed the argument \texttt{--all}.
I set the error rate using the parameter \texttt{-e}.
In paired-end mode, the parameters used were \texttt{--library-length INS --library-error ERR}.
The number of threads was set with the parameter \texttt{-t}.

\paragraph{GEM}
Version 1.376 was used.
I set the error rate using the parameters \texttt{-m} and \texttt{-e}, then I disabled adaptive mapping using the parameter \texttt{--quality-format ignore}.
In best-mapping, to analyze only the best stratum, I passed the argument \texttt{-s 0};
in all-mapping, to analyze all strata, I passed \texttt{-d all -D all -s all --max-big-indel-length 0}.
In single-end mode, I passed the parameter \texttt{--expect-single-end-reads}; in paired-end mode, I passed \texttt{--paired-end-alignment}, along with \texttt{--min-insert-size MIN --max-insert-size MAX}, and \texttt{--map-both-ends} to select the workflow mapping both reads independently.
The number of threads was selected using the parameter \texttt{-t}.

\paragraph{Bowtie\,2}
Version 2.2.1 was used.
I used the parameter \texttt{--end-to-end} to enforce semi-global read alignments.
%For the Rabema benchmark I used the parameter \texttt{-k 100}.
In paired-end mode, I used the parameters \texttt{--minins MIN --maxins MAX}.
The number of threads was selected using the parameter \texttt{-p}.

\paragraph{BWA}
Version 0.7.7-r441 was used.
%For the Rabema experiment I passed the parameter \texttt{-N} to \texttt{aln} and \texttt{-n 100} to \texttt{samse}.
I used the parameter \texttt{-t} to select the number of threads in the \texttt{aln} step;
the \texttt{sampe} and \texttt{samse} steps were performed using one thread since BWA does not offer any parallelization here.

\paragraph{Hobbes\,2}
Version 2.0 was used.
I built the index using the recommended $q$-gram length 11.
I enabled edit distance with \texttt{--indels} and set the distance threshold using the parameter \texttt{-v}.
In paired-end mode, I used the parameters \texttt{--pe --min MIN --max MAX}.
Multi-threading was enabled using \texttt{-p}.

\paragraph{RazerS\,3}
Version 3.3 was used.
I set the error rate through the parameter \texttt{-i}, \eg \texttt{-i 95} to map within an error rate of 5\,\%.
I passed the option \texttt{-rr 100} to set the recognition rate to 100\,\% and \texttt{-m 1000000} to output all mapping locations per read.
In paired-end mode, the parameters used were \texttt{--library-length INS --library-error ERR}.
The number of threads was set with the \texttt{-tc} parameter.
